---
title: "Forecasting of Family Planning Commodities in Ethiopia"
format:
  revealjs: 
    theme: moon
editor: visual
---

## Why ?

-   5.3 million children under five predicted to die annually, with half in sub-Saharan Africa.

-   The maternal mortality rates remain extremely high, with two-thirds of global deaths occurring in sub-Saharan Africa.

-   Nigeria contributes around 20% of all maternal fatalities worldwide.

-   Adequate family planning is linked to a reduction in infant mortality
    rates, as it promotes healthier pregnancies and better child-rearing
    practices.

## **Ethiopia's Commitment to Family Planning**

-   Family planning recognized as a human right in Ethiopia's Constitution (Article 35(9)).

-   Government abolished contraceptive tax in 2007 and enhanced domestic financial resources for the FP program.

-   Annual allocation to the FP budget has increased over the last decade.

-   Despite progress, the country's FP program heavily relies on external funding.

## Challenges 

-   The Data collection was poor in some districts, affecting the methodologies that could be used.

-   The COVID-19 pandemic created a local trend which meant we needed strictly positive forecasting.

-   The scale of data for each medicine and number of regions.

## Oxytocin

-   Oxytocin's role in family planning is of global health significance, contributing to safe childbirth and reproductive health.

-   Oxytocin is a hormone that plays a crucial role in stimulating uterine contractions during labor and childbirth.

-   Administration of oxytocin is a common practice to prevent postpartum hemorrhage (excessive bleeding) after childbirth.

## Initial Data Analysis

```{r message=FALSE, warning=FALSE, echo=FALSE}

library(tidyverse)
library(fpp3)

data <- read_csv(here::here("data/famillyplanning_maternal_child_health_products.csv"))

data$Date <- as.Date(paste0(data$month, "01"), format = "%Y %b %d")

tsibble_data <- data %>%
  as_tsibble(index = Date, key = c(region, item)) %>%
  mutate(Date = yearmonth(Date)) %>%
  select(Date, region, item, total_issued)

data_Oxytocin <- tsibble_data %>% 
  filter(item == "Oxytocin - 10 Units/ml in 1ml Ampoule - Injection" 
         )

data_Oxytocin%>% 
  autoplot() +
  facet_wrap(~ region, scales = "free") +
  theme(legend.position = "none") +
  labs(x = "Year", y = "Total Issued", title = "Initial Data Analysis for Oxytocin") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  
```

## Forecasting in Addis Ababa using ARIMA

```{r  message=FALSE, warning=FALSE, echo=FALSE}

data_region_oxy <- data_Oxytocin %>% filter(year(Date) < 2022 | (year(Date) == 2022 & month(Date) <= "May"))
  
  # Model creation
  models_region_oxy <- data_region_oxy %>% 
    model(
      ARIMA = ARIMA(total_issued),
      ETS = ETS(total_issued),
      naive = NAIVE(total_issued)
    )
  # Forecasting
  forecast_region_oxy <- models_region_oxy %>% 
    forecast(h = "1 years")
  
  # Ensure non-negative forecasts
  forecast_pos_region_oxy <- forecast_region_oxy %>%
    mutate(.mean = ifelse(.mean < 0, 0, .mean))
```

```{r }

# Forecasting Using ARIMA model

forecast_pos_region_oxy %>% 
  filter(.model == "ARIMA" ) %>%
  filter(region == "Addis Ababa") %>% 
   dplyr::mutate(value = distributional::dist_truncated(.mean, 0)) %>% 
  autoplot(data_Oxytocin %>% 
             filter(region == "Addis Ababa")) +
  labs(x = "Year", y = "Total Issued", title = "Forecasting using ARIMA model")
```

## Forecasting in Addis Ababa using ETS

```{r}
# Forecasting Using ARIMA model

forecast_pos_region_oxy %>% 
  filter(.model == "ETS" ) %>%
  filter(region == "Addis Ababa") %>% 
   dplyr::mutate(value = distributional::dist_truncated(.mean, 0)) %>% 
  autoplot(data_Oxytocin %>% 
             filter(region == "Addis Ababa")) +
  labs(x = "Year", y = "Total Issued", title = "Forecasting using ETS model")

```

## Forecasting in Addis Ababa using NAIVE

```{r}
forecast_pos_region_oxy %>% 
  filter(.model == "naive" ) %>%
  filter(region == "Addis Ababa") %>% 
   dplyr::mutate(value = distributional::dist_truncated(.mean, 0)) %>% 
  autoplot(data_Oxytocin %>% 
             filter(region == "Addis Ababa")) +
  labs(x = "Year", y = "Total Issued", title = "Forecasting using NAIVE model")
```

## Selecting the Appropriate model

```{r}
 # Cross-validation results
  
  # Stretch tsibble for cross-validation
  stretched_data_region_oxy <- stretch_tsibble(data_Oxytocin, .init = 1, .step = 1)

  cv_results_region_oxy <- stretched_data_region_oxy %>%
    filter(region == "Addis Ababa") %>% 
    model(
      ARIMA = ARIMA(total_issued),
      ETS = ETS(total_issued),
      naive = NAIVE(total_issued)
    ) %>%
    forecast(h = "1 years") %>%
    fabletools::accuracy(data_Oxytocin)
  
  kableExtra::kable(cv_results_region_oxy %>% 
                      select(-item,
                             -MAPE,
                             -MPE))
```

## Winkler Scores

```{r}
winkler_scores <- forecast_pos_region_oxy %>% 
  filter(region == "Addis Ababa") %>% 
  fabletools::accuracy(data_Oxytocin,
                       list(winkler = winkler_score), level = 80)

kableExtra::kable(winkler_scores)
```

## Final Thoughts

-   Address issues in data collection methodologies in certain districts.

-   ETS model showcased promising performance, but a comprehensive evaluation is crucial.
